---
title: "Understanding functions in R"
output: 
  html_document:
#    code_folding: hide
  toc: true
highlight: pygments
---
  
This is a quick guide to understanding the behaviour of functions in R. 


```{r message=FALSE}
library(dplyr)
```


<br><br>
Functions will only print the last line of output. Functions can only return one object (but it can be a list of other objects), and the function will return the result of the last executed line of code by default. If `return()` is used, the function will stop running at that point (so any code inside the function below the `return()` will not be executed). Note that if the last statement happens to be a print statement, the function will 'return' that (and only that, meaning any object you thought may also be returned will not be returned)


```{r}

ftest <- function() {
  2 * 2
  3 * 3
}

ftest()


```

<br><br>

If we call a function that operates on some variable, then the variable must be available in the environment in which the function is called, or some ancestor environment.

The function will first check it's local environment before checking the global environment:


```{r}


ftest <- function() {
  y <- 2.3
  y * 2
}
ftest()

```


<br><br>


Here we do not define y within the scope of the function, so R looks to the parent, and finds y in the global environment

```{r}
y <- 5

ftest <- function() { y * 2 }
ftest()



```


<br>

If y is not available in the environment from which the function is called or any ancestors, it errors

```{r, eval=FALSE, echo=TRUE}

rm(y)
ftest()
# Error in ftest() : object 'y' not found

```


<br><br>

If we tell a function that it will have arguments passed to it, then it must receive those arguments *unless* the function doesn't actually use them
Also, it doesn't matter whether something of the same name is available globally, if the argument is not provided in the function call, it will error

```{r, eval=FALSE, echo=TRUE}

ftest <- function(y) { y * 2 }
ftest()
# Error in ftest() : argument "y" is missing, with no default

```

<br><br>

Function doesn't receive y or z, but it doesn't use them so there is no error

```{r}

ftest <- function(y, z) { 2 * 2 }
ftest()


```

<br><br>

```{r eval=FALSE, echo=TRUE}
ftest <- function() { z * 2 }
ftest()
```
<br>

```{r, eval=FALSE, echo=TRUE}
# However, lazy evaluation should have meant this wouldn't error 
ftest <- function(y, x) { y * 2 }
ftest()
```

<br>
```{r, eval=FALSE, echo=TRUE}
# Here we 
ftest <- function(y) { y * 2 }
ftest(2)
# [1] 4

```




<br><br><br>

## Understanding error: "promise already under evaluation: recursive default argument reference or earlier problems?"
From [here](https://stackoverflow.com/questions/8425348/unexpected-behaviour-with-argument-defaults)

This is simply because R won't allow you to use the same argument name for a function within a function


```{r, eval=FALSE, echo=TRUE}

y <- 5

# f knows what x is because y is defined in the global environment
f <- function(x=y) x^2
f() # Works fine

# f2 
f2 <- function(y=y) y^2
f2() # Errors


```

<br><br>


## Understanding functions within functions 


Here we can see that the inner_function can access the arguments that were provided to the outer_function,
Which makes sense, since functions can access objects from the global environment
Possible lesson: environments can access all the variables from their ancestors

```{r, eval=FALSE, echo=TRUE}

outer_function <- function(x, y, ...) { 
  
  
  inner_function <- function(x=x, y=y) {
    
    print(x)
    print(y)
    
    
    args <- list(...)
    print(args)
    print(args[[1]])
    print(args[[2]])
  }
  
  inner_function()
  
  }

outer_function("hi", "bye", "here", "now")


```
<br>
```{r}


HelloWorld <- function(...) {
  arguments <- list(...)
  paste(arguments)
  arguments[[1]] %>% print(.)
  arguments[[2]] %>% print(.)
}

HelloWorld("Hello", "World", "!")


```

<br>
```{r}

HelloWorld <- function(x, ...) {
  arguments <- list(...)
  paste(arguments)
  arguments[[1]] %>% print(.)
  arguments[[2]] %>% print(.)
}

HelloWorld("Hello", "World", "!")

```

<br>

```{r}


HelloWorld <- function(x, ...) {
  arguments <- list(...)
  paste(arguments)
  x %>% print(.)
  arguments[[1]] %>% print(.)
  arguments[[2]] %>% print(.)
}

HelloWorld("Hello", "World", "!")

```


```{r}



HelloWorld <- function(x, ...) {
  arguments <- list(...)
  paste(arguments)
  x %>% print(.)
  arguments[[1]] %>% print(.)
  arguments[[2]] %>% print(.)
  
  
  HelloWorld2 <- function(...) {

   arguments2 <- list(...)
   temp <- arguments2[[1]] %>% paste0(", it works!!")
   return(temp)
  }

  temp <- HelloWorld2(x) # Don't forget to assign what is returned by the inner function, or else you don't be able to access it 

  print(temp)
  
}

# Note in both cases, "world" and "!" are allowed in by the ... (variable number of arguments)
HelloWorld("Hello", "World", "!")
HelloWorld(x = "Hello", "world", "!")

```

<br><br><br>


## Using the ellipses to pass unnamed arguments to a function 

```{r} 

do_things <- function(weather, show_weather, ...) {
  
  if (show_weather) { paste0("It's ", weather, " outside!") %>% print }
  
  
  arguments <- list(...)
  
  paste0("I like the numbers ", arguments[[1]], ", ", arguments[[2]], ", and ", arguments[[3]], ".") %>% print()
  { arguments[[4]] + arguments[[5]] } %>% print()
  arguments[[6]] ^ 2 %>% print()
  
  
}

do_things("hot", TRUE, 1,2,3,4,5,6)

```


<br><br><br>




## Returning console output 

### Example 1

This function will print the console output for the last line (only) that's executed inside the function

```{r}


HelloWorld <- function(...) {
  arguments <- list(...)
  paste(arguments)
  arguments[[1]]
  arguments[[2]] # Try commenting this line out 
  arguments[[3]] # Try commenting this line out
}

HelloWorld("Hello", "World", "!")





```





### Example 2

In this function, nothing will be seen in the console

```{r} 

add_me <- function(...) {
  
  arguments <- list(...)
  
  for (i in 1:length(arguments)) {
    arguments[i]
  }
}

add_me(1,2,3,4,5,6)


```

But this variation print everything we explicity tell it to with print() (but try commenting that line out and simply having arguments[i])


```{r} 

add_me <- function(...) {
  
  arguments <- list(...)
  
  for (i in 1:length(arguments)) {
    print(arguments[i])
    # arguments[i]
  }
}

add_me(1,2,3,4,5,6)


```





One elegant solution would be to pipe everything to print()


```{r}

add_me <- function(...) {
  
  arguments <- list(...)
  
  for (i in 1:length(arguments)) {
    arguments[i] %>% print()
  }
  
  
}

add_me(1,2,3,4,5,6)

```









```{r, eval=FALSE, echo=TRUE}

# This works
outer_function <- function(x) { 
  inner_function <- function(x) {
    
    return(paste0(x, ", it worked!"))
  }
  
  output_from_inner_function <- inner_function(x)
  print(output_from_inner_function)
}
outer_function("hi")








outer_function <- function(x, ...) {
  args1 <- list(...)
  inner_function <- function(x, args1) {
    args2 <- list(...)
    print(paste(args2))
  }
}

outer_function("hi", "bye")



# Can we somehow print the contents of the inner function?
# ANS: Yes, just store the output (or assigned variables) available and print them
outer_function <- function() {
  inner_function <- function() {
    "hi"
  }
  toprint <- inner_function()
  print(toprint)
}
outer_function()


```






```{r}

#----- Example 1 -----#
# calling an object's name in a function is no different to doing so normally, that is, 
# it will simply return that object's value

fun <- function(x) { x }
fun("hi")
# [1] "hi"





#----- Example 2 -----#
# Function calls inside a function will return the same output as though that code had been called normally within the outer function
# BUT it will only return the *last line* of that output - why? I'm not sure! But if you want more than just the last line, assign to a list and return it!
outer_function <- function() {
  
  inner_function <- function() {
    "hi"
    parent.frame()
  } 
  
  inner_function() 
}

outer_function()


```










# Notes on Environments in R


```{r, eval=FALSE, echo=TRUE}


globalenv()
globalenv() %>% parent.env # The parent environment of the global environment is the last package opened (who knows why?) - unless no packages are open in which case it doesn't have a parent env
globalenv() %>% parent.env %>% parent.env



#---- Create a nested environment -----#
# From: https://stackoverflow.com/questions/10904124/global-and-local-variables-in-r


test.env <- new.env()

assign('abcd', 100, envir=test.env)
# or simply
test.env$abcd <- 100

get('abcd') # abcd cannot be found since it is not defined in this environment
get('abcd', envir=test.env) # now it can be found




#----- Observe the objects inside an environment -----#
# From here: https://stackoverflow.com/questions/8142941/how-to-search-an-environment-using-ls-inside-a-function
# QQQ why .GlobalEnv below? rather than R_GlobalEnv, that is, the output from parent.frame() ?

x <- 1:10
f <- function() ls()
g <- function() ls(envir=.GlobalEnv)
h <- function() ls(envir=.GlobalEnv, pattern="[fg]")
f()
# character(0)
g()
# [1] "f" "g" "h" "x"
h()
# [1] "f" "g"



f <- function() {
  output <- list()
  output[[1]] <- ls()
  output[[2]] <- parent.frame()
  return(output)
}


f()



f <- function() { parent.frame() }
f()






# Four special environments 

globalenv() # Environment in which you normally work
baseenv() # Base environment
emptyenv() # The ultimate ancestor of all environments 
environment() # Returns the current environment
identical(globalenv(), environment()) # [1] TRUE - which makes sense

search() # lists all parents of the global environment 


```





## Behaviour when setting default arguments

A simple function that will double a number..

```{r}

double_me <- function(x) { 2 * x }
double_me(12.436)

```


### How to set a default value for a function

Here a function will double x unless triple = TRUE, in which case it will triple it

```{r}
multiply_me <- function(x, triple) { 
  
  if (missing(triple)) { triple <- FALSE}
  
  if (triple == TRUE) {
  3 * x 
    } else {
      2 * x
      } %>% return(.)

}

multiply_me(123.234)
multiply_me(123.234, triple = TRUE)







```



# Futher notes on environments, scopes, and assignments

### We can create a variable in a specific environment in two ways
```{r}

# https://stackoverflow.com/questions/54830589/strictly-speaking-does-the-scoping-assignment-assign-to-the-parent-environme

env = new.env()
env$abc <- 2222 


local(abc <- 2222, env)


# E.g.
# Create environments 
rm(list=ls()); env = new.env(); child_of_env = new.env(parent = env)
# Note, a variable can be assigned to a particular environment like so
env$abc <- 2222
ls(env)
# [1] "abc"

```

### Three experiments to show: 
    <ul>
      <li>How to make an environment inside another environment</li>
      <li>How to assign to a child environment</li>
      <li>How using x <<- 2 in a child environment will walk up environments until it finds x and assign it, 
      if it doesn't find x in any higher environment, it will assign it in the global environment</li>
    </ul>


```{r}


# Create environments 
rm(list=ls()); env = new.env(); child_of_env = new.env(parent = env)

local(x <- 42, child_of_env)
ls(env)
ls(child_of_env)
ls()

# Create environments 
rm(list=ls()); env = new.env(); child_of_env = new.env(parent = env)

local(x <<- 42, child_of_env)
ls(env)
ls(child_of_env)
ls()




# Create environments
rm(list=ls()); env = new.env(); child_of_env = new.env(parent = env)

env$x <- 2
local(x <<- 42, child_of_env)
ls(env)
env$x


```
















